---
title: Quantification of snRNA-seq data after sleep deprivation with salmon 
author: Davide Risso
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Many thanks to Stephanie Hicks for providing code and advice on how to run salmon on single-nucleus data.

```{r}
library(here)
```

We create a folder to store reference files (e.g. genome and cDNA) and quantification files (quantified counts) from salmon.  

```{r}
if(!dir.exists(here("salmon_files"))){
  dir.create(here("salmon_files"))
}

if(!dir.exists(here("salmon_quants"))){
  dir.create(here("salmon_quants"))
}
```

# Quantification 

## Download files needed for salmon

We need to download 3 files:

1. `GRCm38.primary_assembly.genome.fa.gz` - nucleotide (DNA) sequences of the **GRCH38 primary genome assembly** (chromosomes and scaffolds -- i.e. unplaced scaffolds?)
2. `gencode.vM24.transcripts.fa.gz` - nucleotide (DNA) sequences of **all transcripts** on reference chromosomes
3. `gencode.vM24.annotation.gtf.gz` - gene annotation on the reference chromosomes (i.e. for humans, these are chromosomes 1 to 22, X, and Y), i.e. locations of genes and other information about the genes, gene structure
  * Gene transfer format (GTF) is a file format used to hold information about gene structure. It is a tab-delimited text format based on the general feature format (GFF), but contains some additional conventions specific to gene information.

Source: https://www.gencodegenes.org/human/

```{r}
# download GENCODE primary assembly fasta file
if(!file.exists(here("salmon_files", "GRCm38.primary_assembly.genome.fa.gz"))){
  tar_gz_file <- "ftp://ftp.ebi.ac.uk/pub/databases/gencode/Gencode_mouse/release_M24/GRCm38.primary_assembly.genome.fa.gz"
  download.file(tar_gz_file, 
                destfile = here("salmon_files", "GRCm38.primary_assembly.genome.fa.gz"), 
                method = "wget")
}

# download GENCODE transcripts fasta file
if(!file.exists(here("salmon_files", "gencode.vM24.transcripts.fa.gz"))){
  tar_gz_file <- "ftp://ftp.ebi.ac.uk/pub/databases/gencode/Gencode_mouse/release_M24/gencode.vM24.transcripts.fa.gz"
  download.file(tar_gz_file, 
                destfile = here("salmon_files", "gencode.vM24.transcripts.fa.gz"), 
                method = "wget")
}

# download GENCODE gtf file
if(!file.exists(here("salmon_files", "gencode.vM24.annotation.gtf.gz"))){
  tar_gz_file <- "ftp://ftp.ebi.ac.uk/pub/databases/gencode/Gencode_mouse/release_M24/gencode.vM24.annotation.gtf.gz"
  download.file(tar_gz_file, 
                destfile = here("salmon_files", "gencode.vM24.annotation.gtf.gz"), 
                method = "wget")
}
```


## Make pre-mRNA files for salmon

We need to create a set of pre-mRNA sequences to save to a FASTA file. 
This is needed because we have single nucleus RNA-seq data so we need to take into account the pre-mRNA and mRNA transcripts. 
This code kindly comes from [Charlotte Soneson](https://csoneson.github.io).

```{r}
suppressPackageStartupMessages({
  library(Biostrings)
  library(rtracklayer)
  library(GenomicFeatures)
  library(BSgenome)
})
```

### Combined GTF file

First, make a pre-mRNA GTF file (using the mRNA GTF file). Then combine it with the mRNA GTF file to get a GTF file for both pre-mRNA and mRNA.

```{r}
## Read gtf file and group exons by transcript
ingtf <- here("salmon_files","gencode.vM24.annotation.gtf.gz")
txdb <- GenomicFeatures::makeTxDbFromGFF(ingtf)
grl <- GenomicFeatures::exonsBy(txdb, by = "tx", use.names = TRUE)

## Create the pre-mRNA coordinates by adding all introns to each transcript
grrange <- unlist(range(grl)) # 142552 rows
names(grrange) <- paste(names(grrange), "premrna", sep= ".")

## Save pre-mRNA gtf file
premrna_gtf <- here("salmon_files","gencode.vM24.premrnaannotation.gtf")
rtracklayer::export(grrange, premrna_gtf)

## Combine two gtf files (pre-mRNA and mRNA) to be used in tximeta a bit later
unlist(grl) # 842873 rows
tmp <- c(grrange, unlist(grl)) # 985425 rows, pre-mRNA comes in front of mRNA

## example of a transcript with three exons (mRNA) or entire pre-mRNA region
tmp[grep("ENSMUST00000161581.1", names(tmp)),]

## Save combined pre-mRNA and mRNA gtf file
preandmrna_gtf <- here("salmon_files","gencode.vM24.preandmrnaannotation.gtf")
rtracklayer::export(tmp, preandmrna_gtf)
```

### Combined FASTA file

Now, for each entry in `grrange`, we extract the sequence from the corresponding `seqnames` entry in the `genome`.

**Note**: If you use Ensembl, there will be a discrepancy between the transcripts (e.g. ENST00000456328 vs ENST00000456328.1). 
Technically, it doesn't matter this point, but it will when we merge with the spliced (mRNA) transcripts.
So, if you do this, you will likely need to remove the version numbers before merging the pre-mRNA and mRNA fasta files. 
Or just use the Gencode annotation where the naming is consistent between the transcriptome fasta and the gtf.

```{r}
## Get genomic (DNA) sequence from FASTA file
genomefasta <- here("salmon_files", "GRCm38.primary_assembly.genome.fa.gz")

## Import FASTA file (DNA) with Biostrings::readDNAStringSet
genome <- Biostrings::readDNAStringSet(genomefasta)
names(genome) <- sapply(strsplit(names(genome), " "), .subset, 1) # creates chr1, etc

## Get the sequence of each pre-mRNA molecule
premrnaseq <- BSgenome::getSeq(x = genome, names = grrange)

## Save pre-mRNA sequences to fasta file 
premrna_fasta <- here("salmon_files", "GRCm38.premRNA.fa")
Biostrings::writeXStringSet(premrnaseq, filepath = premrna_fasta)
```

Let's check to make sure the pre-mRNA and the mRNA have the same number of transcripts.

```{bash}
cd salmon_files
zcat gencode.vM24.transcripts.fa.gz | grep ">"  | wc -l # mRNA
grep ">" GRCm38.premRNA.fa | wc -l                     # pre-mRNA
```

Both have the same number of rows (or transcripts): 227462

Now we will put the pre-mRNA (`GRCm38.premRNA.fa`) and mRNA (`gencode.vM24.transcripts.fa.gz`) fasta GENCODE files together. 

```{bash}
cd salmon_files
zcat gencode.vM24.transcripts.fa.gz > gencode.vM24.transcripts.fa
cat gencode.vM24.transcripts.fa GRCm38.premRNA.fa > gencode.vM24.preandmrna.fa
grep ">" gencode.vM24.preandmrna.fa | wc -l
```

Ok there are 285104 (or = 142552*2) transcripts, as we expect. Now let's zip this file up. 
This takes a while (~30-40 mins). 

```{bash}
cd salmon_files
gzip gencode.vM24.preandmrna.fa
```

And clean up our space

```{bash}
cd salmon_files
rm gencode.vM24.transcripts.fa
rm GRCm38.primary_assembly.genome.fa.gz
```

We have made a GTF file (`gencode.vM24.preandmrnaaannotation.gtf`) and a FASTA file (`gencode.vM24.preandmrna.fa`) for the combined pre-mRNA and mRNA.

## Build salmon index 

The `-t` argument is the input transcripts file. 
The `-i` argument is the index file to create. 
The `--keepDuplicates` argument forces all duplicate transcripts (for example, multiple unspliced transcript of the same gene that are identical for example) that appear in the input will be retained and quantified separately. 
If you keep the duplicates they will be assigned identical expression levels since salmon can’t tell them apart. 
When you aggregate on the gene level, this will not make a difference any more. 
Therefore, I do not keep the duplicates as we are interested in gene level aggregation. 
The `--gencode` flag will handle the composite fasta headers in GENCODE transcript fasta files and  split the transcript name at the first '|' character. 
The `--threads` argument says how many threads to use when building the index. 

```{bash}
salmon index -t salmon_files/gencode.vM24.preandmrna.fa.gz -i salmon_files/gencode.vM24_salmon-index-v1.0.0-mRNA --gencode --threads 20
```

We also build an index with only the mRNA transcripts.

```{bash}
salmon index -t salmon_files/gencode.vM24.transcripts.fa.gz -i salmon_files/gencode.vM24_salmon-index-v1.0.0-mRNAonly --gencode --threads 20
```

And with only the pre-mRNA transcripts.

```{bash}
salmon index -t salmon_files/GRCm38.premRNA.fa.gz -i salmon_files/gencode.vM24_salmon-index-v1.0.0-premRNA --gencode --threads 20
```


## Create transcript-gene map

Alevin needs a transcript-gene map, we can create one starting from the ENSEMBL transcript IDs.

```{r}
ingtf <- here("salmon_files","gencode.vM24.annotation.gtf.gz")
txdb <- GenomicFeatures::makeTxDbFromGFF(ingtf)
tx <- transcripts(txdb, columns=c("tx_name", "gene_id"))
txdf <- as.data.frame(mcols(tx))
txdf2 <- txdf
txdf2$tx_name <- paste0(txdf2$tx_name, ".premrna")
txdfall <- rbind(txdf, txdf2)
write.table(txdfall, row.names = FALSE, quote = FALSE, col.names = FALSE,
            file = here("salmon_files", "tx_gene_map.tsv"),
            sep = "\t")
```

We do the same with mRNA only.

```{r}
ingtf <- here("salmon_files","gencode.vM24.annotation.gtf.gz")
txdb <- GenomicFeatures::makeTxDbFromGFF(ingtf)
tx <- transcripts(txdb, columns=c("tx_name", "gene_id"))
txdf <- as.data.frame(mcols(tx))
write.table(txdf, row.names = FALSE, quote = FALSE, col.names = FALSE,
            file = here("salmon_files", "tx_gene_map_mrna.tsv"),
            sep = "\t")
```

And with pre-mRNA only.

```{r}
ingtf <- here("salmon_files","gencode.vM24.annotation.gtf.gz")
txdb <- GenomicFeatures::makeTxDbFromGFF(ingtf)
tx <- transcripts(txdb, columns=c("tx_name", "gene_id"))
txdf <- as.data.frame(mcols(tx))
txdf$tx_name <- paste0(txdf$tx_name, ".premrna")
write.table(txdf, row.names = FALSE, quote = FALSE, col.names = FALSE,
            file = here("salmon_files", "tx_gene_map_premrna.tsv"),
            sep = "\t")
```

## Run alevin 

```{bash}
mkdir salmon_quants_preandmrna_1C
mkdir salmon_quants_preandmrna_2E
mkdir salmon_quants_preandmrna_3C
mkdir salmon_quants_preandmrna_4E
mkdir salmon_quants_preandmrna_5C
mkdir salmon_quants_preandmrna_8E

salmon alevin -l ISR -1 /home/risso/data/peixoto_2020-01/1C/1C_S1_L001_R1_001.fastq.gz /home/risso/data/peixoto_2020-01/1C/1C_S1_L002_R1_001.fastq.gz /home/risso/data/peixoto_2020-01/1C/1C_S1_L003_R1_001.fastq.gz /home/risso/data/peixoto_2020-01/1C/1C_S1_L004_R1_001.fastq.gz -2 /home/risso/data/peixoto_2020-01/1C/1C_S1_L001_R2_001.fastq.gz /home/risso/data/peixoto_2020-01/1C/1C_S1_L002_R2_001.fastq.gz /home/risso/data/peixoto_2020-01/1C/1C_S1_L003_R2_001.fastq.gz /home/risso/data/peixoto_2020-01/1C/1C_S1_L004_R2_001.fastq.gz --chromiumV3 -i salmon_files/gencode.vM24_salmon-index-v1.0.0-mRNA -p 20 -o salmon_quants_preandmrna_1C --tgMap salmon_files/tx_gene_map.tsv --forceCells 10000

salmon alevin -l ISR -1 /home/risso/data/peixoto_2020-01/2E/2E_S1_L001_R1_001.fastq.gz /home/risso/data/peixoto_2020-01/2E/2E_S1_L002_R1_001.fastq.gz /home/risso/data/peixoto_2020-01/2E/2E_S1_L003_R1_001.fastq.gz /home/risso/data/peixoto_2020-01/2E/2E_S1_L004_R1_001.fastq.gz -2 /home/risso/data/peixoto_2020-01/2E/2E_S1_L001_R2_001.fastq.gz /home/risso/data/peixoto_2020-01/2E/2E_S1_L002_R2_001.fastq.gz /home/risso/data/peixoto_2020-01/2E/2E_S1_L003_R2_001.fastq.gz /home/risso/data/peixoto_2020-01/2E/2E_S1_L004_R2_001.fastq.gz --chromiumV3 -i salmon_files/gencode.vM24_salmon-index-v1.0.0-mRNA -p 20 -o salmon_quants_preandmrna_2E --tgMap salmon_files/tx_gene_map.tsv --forceCells 10000

salmon alevin -l ISR -1 /home/risso/data/peixoto_2020-01/3C/3C_S1_L001_R1_001.fastq.gz /home/risso/data/peixoto_2020-01/3C/3C_S1_L002_R1_001.fastq.gz /home/risso/data/peixoto_2020-01/3C/3C_S1_L003_R1_001.fastq.gz /home/risso/data/peixoto_2020-01/3C/3C_S1_L004_R1_001.fastq.gz -2 /home/risso/data/peixoto_2020-01/3C/3C_S1_L001_R2_001.fastq.gz /home/risso/data/peixoto_2020-01/3C/3C_S1_L002_R2_001.fastq.gz /home/risso/data/peixoto_2020-01/3C/3C_S1_L003_R2_001.fastq.gz /home/risso/data/peixoto_2020-01/3C/3C_S1_L004_R2_001.fastq.gz --chromiumV3 -i salmon_files/gencode.vM24_salmon-index-v1.0.0-mRNA -p 20 -o salmon_quants_preandmrna_3C --tgMap salmon_files/tx_gene_map.tsv --forceCells 10000

salmon alevin -l ISR -1 /home/risso/data/peixoto_2020-01/4E/4E_S1_L001_R1_001.fastq.gz /home/risso/data/peixoto_2020-01/4E/4E_S1_L002_R1_001.fastq.gz /home/risso/data/peixoto_2020-01/4E/4E_S1_L003_R1_001.fastq.gz /home/risso/data/peixoto_2020-01/4E/4E_S1_L004_R1_001.fastq.gz -2 /home/risso/data/peixoto_2020-01/4E/4E_S1_L001_R2_001.fastq.gz /home/risso/data/peixoto_2020-01/4E/4E_S1_L002_R2_001.fastq.gz /home/risso/data/peixoto_2020-01/4E/4E_S1_L003_R2_001.fastq.gz /home/risso/data/peixoto_2020-01/4E/4E_S1_L004_R2_001.fastq.gz --chromiumV3 -i salmon_files/gencode.vM24_salmon-index-v1.0.0-mRNA -p 20 -o salmon_quants_preandmrna_4E --tgMap salmon_files/tx_gene_map.tsv --forceCells 10000

salmon alevin -l ISR -1 /home/risso/data/peixoto_2020-01/5C/5C_S1_L001_R1_001.fastq.gz /home/risso/data/peixoto_2020-01/5C/5C_S1_L002_R1_001.fastq.gz /home/risso/data/peixoto_2020-01/5C/5C_S1_L003_R1_001.fastq.gz /home/risso/data/peixoto_2020-01/5C/5C_S1_L004_R1_001.fastq.gz -2 /home/risso/data/peixoto_2020-01/5C/5C_S1_L001_R2_001.fastq.gz /home/risso/data/peixoto_2020-01/5C/5C_S1_L002_R2_001.fastq.gz /home/risso/data/peixoto_2020-01/5C/5C_S1_L003_R2_001.fastq.gz /home/risso/data/peixoto_2020-01/5C/5C_S1_L004_R2_001.fastq.gz --chromiumV3 -i salmon_files/gencode.vM24_salmon-index-v1.0.0-mRNA -p 20 -o salmon_quants_preandmrna_5C --tgMap salmon_files/tx_gene_map.tsv --forceCells 10000

salmon alevin -l ISR -1 /home/risso/data/peixoto_2020-01/8E/8E_S1_L001_R1_001.fastq.gz /home/risso/data/peixoto_2020-01/8E/8E_S1_L002_R1_001.fastq.gz /home/risso/data/peixoto_2020-01/8E/8E_S1_L003_R1_001.fastq.gz /home/risso/data/peixoto_2020-01/8E/8E_S1_L004_R1_001.fastq.gz -2 /home/risso/data/peixoto_2020-01/8E/8E_S1_L001_R2_001.fastq.gz /home/risso/data/peixoto_2020-01/8E/8E_S1_L002_R2_001.fastq.gz /home/risso/data/peixoto_2020-01/8E/8E_S1_L003_R2_001.fastq.gz /home/risso/data/peixoto_2020-01/8E/8E_S1_L004_R2_001.fastq.gz --chromiumV3 -i salmon_files/gencode.vM24_salmon-index-v1.0.0-mRNA -p 20 -o salmon_quants_preandmrna_8E --tgMap salmon_files/tx_gene_map.tsv --forceCells 10000
```

## Set up the tximeta package 

Next, we use the `tximeta` package to create a SummarizedExperiment object.
If we try to run `tximeta(coldata)`, that will fail because `tximeta` won't be able to detect automatically what the reference transcriptome is. 
So, here we create our own [linked transcriptome](https://bioconductor.org/packages/release/bioc/vignettes/tximeta/inst/doc/tximeta.html#linked_transcriptomes) from the combined GTF file.

A `linkedTxome` records key information about the sources of the transcript FASTA files, and the location of the relevant GTF file. 
It also records the checksum of the transcriptome that was computed by Salmon during the index step.

For some reason this does not work

```{r, eval=FALSE}
suppressPackageStartupMessages({
  library(tximeta)
})

# create linkedTranscriptome 
index_dir <- here("salmon_files", "gencode.vM24_salmon-index-v1.0.0-mRNA")
fasta_path <-  here("salmon_files", "gencode.vM24.preandmrna.fa.gz")
gtf_path <-  here("salmon_files", "gencode.vM24.preandmrnaaannotation.gtf") 

json_file <- here("salmon_files", paste0(basename(index_dir), ".json"))
makeLinkedTxome(indexDir=index_dir,
                source="other", organism="Mus musculus",
                release="other", genome="GRCm38",
                fasta=fasta_path, gtf=gtf_path,
                jsonFile=json_file) # this command will add the index to the cache automatically
```

## Create SummarizedExperiment object

```{r}
library(tximeta)
library(scater)
library(scran)

alevin_quant <- here("salmon_quants_preandmrna_1C/alevin", "quants_mat.gz")
alevin <- tximeta::tximeta(coldata = data.frame(files = alevin_quant,
                                                names = "Sample1C"), 
                           type = "alevin")
whitelist <- read.table(here("salmon_quants_preandmrna_1C/alevin", "whitelist.txt"),
                        stringsAsFactors = FALSE)
alevin$whitelist <- colnames(alevin) %in% whitelist[,1]
alevin <- as(alevin, "SingleCellExperiment")
saveRDS(alevin, file = here("salmon_quants_preandmrna_1C", "sce.rds"))
```

